A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-31 paper.tex
Create Ancestral graphs: history section
- Adapt most of content from the Phylogenetic networks section to this
new section or a discussion point.

% \section*{Ancestral graphs and stochastic processes}
% The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
% hudson1983testing, tajima1983evolutionary}
% provides the theoretical underpinning for much of contemporary population genetics.
% \citet{hudson1983properties,hudson1990gene} extended the coalescent to include
% recombination
% and the model has proved to be an especially powerful basis for
% simulations~\citep{hudson2002generating,baumdicker2021efficient}.
% \cite{kelleher2016efficient} provide a description of his algorithm along with
% an efficient implementation, the details of which turn out to be highly relevant to
% later sections of the current paper (see the ``\nameref{Ancestry_resolution}'' section).

@@

% Similar graph encodings have also been introduced to model other mechanisms
% which change the local ancestral tree along the genome, such as gene
% conversion~\citep{wiuf2000coalescent} and
% horizontal gene transfer~\citep{baumdicker2014infinitely}.
% [Mention the
% SMC~\citep{mcvean2005approximating,marjoram2006fast}
% and ClonalFrame~\citep{didelot2007inference} approximations somewhere.]

@@

Another limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts? Gregor: I recall this paper from
Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

@@

@@

\item We should discuss other types of ancestry network also.
The crucial distiction between an ARG and a phylogenetic network (I think?) is
the presence of position information specifying the location of
recombination events (and by extension, the complete passage
of ancestral material the graph). Without position information
we know that a particular node was an ancestor somewhere along the
sequence, but not exactly where.

\item We make the point about expressivity of the eARG pretty
hard, but we don't mention anything about indels and structural
rearrangements etc. It's probably worth mentioning genome graphs
somewhere just to fend off this criticism. The high-level point
would be that it's unlikely to be easier to model this stuff
using node annotations.

@@

\textcolor{red}{Discussion below on phylogenetic networks: we need to
reduce the text below down to one paragraph, which
defines the relationship between what we define as an ARG and
what people mean by phylogenetic networks from the literature.
We probably don't need to get into quite so much detail on
the different types that people have studies. Essentially,
we're ending the section with a brief review which situates
our definition in the literature. I think pulling apart the
idea of a ``reticulation'' vs what we call a recombinant
is essential here.}

The term \emph{phylogenetic network} has been broadly accepted to describe the most general
type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
that labels some of its nodes'' (although more narrow and context-dependent definitions can also
be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.

Some sub-classes of explicit rooted phylogenetic networks are defined based on the evolutionary event that the
reticulation nodes capture \citep{huson2010phylogenetic}. In \emph{reassortment} networks, used to
model the evolution of viruses, reticulation nodes correspond to exchange of
(non-recombining) genomic segments between viral particles inside a co-infected host.
In \emph{hybridisation} networks, reticulation nodes correspond to genetic material from two
different species combining to create a hybrid. In \emph{recombination} networks, the reticulation nodes
correspond to recombination events (with leaves labelled by sequences and edges labelled by mutations).
The ARG is a subtype of a recombination network under this definition (restricting to one crossover
recombination per reticulation node).
% AI: Not sure about this - do ARGs also encompass multiple crossover recombinations and gene
% conversion (seems implied above)?

Sub-classes of explicit rooted phylogenetic networks can also be defined by restrictions on their topology.
In \emph{binary networks}, all internal nodes must have either in-degree two and out-degree one or in-degree
one and out-degree two (with the root having out-degree two and the leaves in-degree 1) \citep{steel2016phylogeny}.
In \emph{tree-child networks}, every internal node must have at least one child that is a tree node;
\emph{tree-sibling networks} are constrained by every reticulation node having at least one sibling that is a
tree node \citep{cardona2008extended}. Further, sub-types can be defined by restrictions on
the topology of subgraphs that contain the reticulation loops (cycles containing a reticulation node).
A reticulation loop that does not share any node with another reticulation loop is termed a gall
\citep[][p.\ 237]{gusfield2014recombinatorics}; a reticulation network containing only galls is
called a \emph{galled tree}. Reticulation loops that are joined together in a reticulation network
by at least one shared node form a blob; a \emph{level-$k$ network} is one where the maximal number
of reticulation loops inside a blob is $k$ \citep{choy2005computing}. Galled trees are level-1 networks,
and the tractability afforded by their relatively simple structure results in a number of attractive
combinatorial properties \citep{wang2001perfect, gusfield2004optimal}. We consider the general case of
an ARG, without applying any of these restrictions.

% AI: have not mentioned any implicit or un-rooted networks (split, median, haplotype, reticulograms, etc).
% Also: are we not defining ARGs as ``binary networks''? Might be good to specify this here.

\textcolor{red}{Discussion on ASG ported from earlier (removed) ``stochastic processes'' section}

The idea of using branching-coalescing stochastic processes
to model genetic ancestral trees has also found application in natural selection.
The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
potential ancestral trees. Weak genic selection is incorporated by sampling a
true ancestry from the ensemble in a non-uniform way.
There is no ASG analogue of the more computationally tractable ``little'' ARG,
though some gains in tractability can be made by considering typed
lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
techniques when recurrent mutation is present \citep{fearnhead2001perfect}.
Extensions of the ASG have been developed to frequency-dependent
selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
loci whereupon branching is due to both selection and
recombination~\citep{donnelly1999genealogical}, and high fecundity
reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.

>>>>

\section*{Ancestral graphs: a brief history}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary} models the ancestry of a sample of
genomes under an idealised population model, and provides the theoretical
underpinning for much of contemporary population genetics.
The coalescent is a stochastic process, where each random realisation
is a genealogical tree describing the genetic
ancestors of the samples and their inheritance relationships.
A great strength of the coalescent as a modelling framework
is its flexibility, and numerous elaborations have been
incorporated~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
In the first extension to the basic process, \cite{hudson1983properties}
analysed the effects of recombination on the random genealogical
trees generated by the coalescent,
producing several fundamental analytical results
as well as describing the elegant algorithm
at the heart of simulation
programs~\citep{hudson2002generating,kelleher2016efficient,baumdicker2021efficient}
used in thousands of studies.
In its original formulation, Hudson's algorithm produced a
sequence of genealogical trees along the genome.
Critically, although the correlation structure among these
local trees was analysed analytically, those correlations
were not captured explicitly: the ancestry of the sample
was viewed as a series of disconnected local trees rather than
as a single cohesive object.

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph
(with particular properties; see section XXX for details),
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details).

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry is regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now broken,
since such parsimonious ARGs are extremely unlikely under the
% JK: I assume this is true; any citations to back it up I wonder?
coalescent with recombination.
\cite{minichiello2006mapping} made this separation of the ARG as a
stochastic process vs the ARG as a data structure explicit, and
for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as examining more general phylogenetic
networks~\citep{huson2010phylogenetic}.

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
However, in parallel with this tendancy to decouple the term
Ancestral Recombination graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model an ensemble of correlated
potential ancestral trees and has been of considerable
and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
A vital distinction between the ARG (process) and ASG, however, is that there
is that there is a hard distinction between the stochastic process
(the graph) and an observable realisation (a single tree) because
the ASG encodes \emph{potential} ancestral trees.
Similarly, the Ancestral Gene Transfer Graph
models horizontal gene transfer~\citep{baumdicker2014infinitely}
by [FINISH ME. Also, any other ASG-like \emph{processes} to mention?].

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form was it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely derived from the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces spurious precision about the location and timing
of recombination events.

@@

@@

Another limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''.
Cite~\citep{wiuf2000coalescent}.
Presumably people
already are in bacterial contexts? Gregor: I recall this paper from
Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

@@

[Mention the
and ClonalFrame~\citep{didelot2007inference} approximation somewhere.]

@@

\item The gARG structure as described here is sufficiently expressive
to model any form of homologous genetic transmission between individuals,
and is far more flexible than the classical eARG approach.
Phylogenetic networks~\citep{huson2010phylogenetic} are a more
general concept, and specific classes capture dynamics that
cannot be modelled using an ARG. For example,
reassortment networks model the exchange of
(non-recombining) genomic segments between viral particles inside a co-infected host,
and hybridisation networks, model the genetic material from
different species combining to create a hybrid.
Similarly Genome Graphs~\cite{novak2017genome} model complicated things
within species. Future work should look to capture all of these things
in a single structure.

@@
